---
title: 'Basic inferential data analysis: tooth growth dataset'
geometry: "left=2cm,right=1cm,top=1cm,bottom=1cm"
output:
  pdf_document: default
  html_document:
    keep_md: yes
---
D.V.

# Synopsis
We're going to analyze the ToothGrowth data in the R datasets package.
Our basic steps are:

- Provide a basic summary of the data.
- Use confidence intervals and/or hypothesis tests to compare tooth growth by supp and dose.
- Make conclusion and the assumptions of analysis.

## Summary of the data
This data frame has 60 observations and 3 variables:


len - Tooth length (numeric).
supp - Supplement type (factor): VC - “vitamin C” or OJ - “orange juice”.
dose - Dose in milligrams (numeric).


Let's see the summary:
```{r}
data(ToothGrowth)
str(ToothGrowth)
#summary(ToothGrowth)
sum <- tapply(ToothGrowth$len, ToothGrowth$supp, FUN=summary)
sum

```


There are a bit of differences in mean and median between supp-groups.
Let's compare them and check for outliers by boxplots.

## Comparing tooth growth by supp and dose
```{r}
boxplot(sum, main = "Bloxplot of tooth length by supplement", 
        xlab = "Supplement",
        ylab = "Tooth length")
library(ggplot2)
ggplot(ToothGrowth, aes(x=factor(dose), y=len, fill=supp))+theme_classic()+
  geom_boxplot()+facet_grid(.~supp)+
  scale_x_discrete("Dose in mg") + scale_y_continuous("Tooth length") + 
  ggtitle("Bloxplot of tooth length for each supplement by dose")

```


Data has a few outliers. In general, median effect of orange juice much larger than C-vitamin. But in higher doses (~2 mg), this effect is almost equal (however, there are unequal variance between groups, so that naive comparing isn't so well).

Let's make a statistical t-tests. The null hypothesis of t-test is that there is no difference in tooth growth between dose and supplement. If we use two sided t-test, so the alternate hypothesis is that the significant difference there is.
```{r}
twotest <- function (x,y){
  with(ToothGrowth, paste("two.sided p-value", 
    round(t.test(x, y, alternative = "two.sided", 
        paired = F, var.equal = F, conf.level = 0.95)$p.value,5)))}
greattest <- function (x,y){
  with(ToothGrowth, paste("greater p-value", 
    round(t.test(x, y, alternative = "greater", 
        paired = F, var.equal = F, conf.level = 0.95)$p.value,5)))}
paste("OJ, VC:", 
      with(ToothGrowth, twotest(len[supp=='OJ'], len[supp=='VC'])),",",
      with(ToothGrowth, greattest(len[supp=='OJ'], len[supp=='VC'])))
paste("OJ, VC | 0.5 mg:", 
      with(ToothGrowth, twotest(len[supp=='OJ' & dose == 0.5], len[supp=='VC'& dose == 0.5])),",",
      with(ToothGrowth, greattest(len[supp=='OJ' & dose == 0.5], len[supp=='VC'& dose == 0.5])))
paste("OJ, VC | 1.0 mg:",
      with(ToothGrowth, twotest(len[supp=='OJ' & dose == 1.0], len[supp=='VC'& dose == 1.0])),",",
      with(ToothGrowth, greattest(len[supp=='OJ' & dose == 1.0], len[supp=='VC'& dose == 1.0])))
paste("OJ, VC | 2.0 mg:",
      with(ToothGrowth, twotest(len[supp=='OJ' & dose == 2.0], len[supp=='VC'& dose == 2.0])))
```


So, there is no significant difference (two sided, at 5% level) between OJ and VS in general and in 2.0 mg dose. The one there is - in 0.5 and 1.0 mg dose. But if we test one sided hypothesis we can confirm that we saw at boxplots: effect of orange juice much larger than C-vitamin in general.


# Conclusions
We saw that vitamin C and orange juice have effect on tooth growth. And vitamin C has much larger effect. Increasing the supplement dose level to 1.0 leads to increased tooth growth in general.


